Estimating species interactions in large, abiotically structured communities with Markov networks and stochastic approximation

### 2015-10-15 Harris_2015/recovery.Rmd
clean up directory

```{r, cache = TRUE, message = FALSE}
library(dplyr)        # For manipulating data structures
library(corpcor)      # For regularized partial covariances
library(rosalia)      # For Markov networks
library(arm)          # For regularized logistic regression
library(BayesComm)    # For joint species distribution modeling
library(RColorBrewer) # For color palette
set.seed(1)
```

@@

```{r, cache = TRUE}
pairs_txt = readLines("fakedata/Pairs.txt")

@@

```{r, cache = TRUE}
fit_all = function(filename){
  ######## Import ########

@@

```{r, cache = TRUE}

@@

```{r, cache = TRUE, message=FALSE}
colors = brewer.pal(8, "Dark2")[c(1, 2, 3, 4, 8, 5, 7)]

@@

```{r, cache = TRUE}
library(ggplot2)

@@

ggplot(z[z$n_sites >0, ], aes(x = `Pairs`, y = truth)) + 
  stat_binhex(bins = 100) + 
  xlab("Estimated coefficient value") + 
  ylab("\"True\" coefficient value") + 
  stat_hline(yintercept = 0, size = 1/8) + 
  stat_vline(xintercept = 0, size = 1/8) + 
  scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
  theme_bw() + 
  ggtitle("Null model (\"Pairs\")") + 
  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$Pairs + 0)))
```

@@

```{r, cache = TRUE}
rep = 12
n = 200
boots = replicate(500,
          {
            boot = lapply(
              sizes, 
              function(n){lapply(
                sample.int(max(z$rep), replace = TRUE),
                function(rep){
                  z[z$rep == rep & z$n_sites == n, ]
                }
              ) %>% bind_rows}
            ) %>% bind_rows

@@

```{r, cache = TRUE}
for(compared in c("partial correlation", "correlation", "GLM", "BayesComm", "partial BayesComm", "Pairs")){
  CI = round(
    quantile(
      boots["Markov network", ] / boots[compared, ], 
      c(.025, .975)), 
    2
  )

@@

>>>>

```{r, cache = TRUE, message = FALSE, eval = FALSE}
library(dplyr)        # For manipulating data structures
library(corpcor)      # For regularized partial covariances
library(rosalia)      # For Markov networks
library(arm)          # For regularized logistic regression
library(BayesComm)    # For joint species distribution modeling
library(RColorBrewer) # For color palette
set.seed(1)
```

@@

```{r, cache = TRUE, eval = FALSE}
pairs_txt = readLines("fakedata/Pairs.txt")

@@

```{r, cache = TRUE, eval = FALSE}
fit_all = function(filename){
  ######## Import ########

@@

```{r, cache = TRUE, eval = FALSE}

@@

```{r, cache = TRUE, message=FALSE, eval = FALSE}
colors = brewer.pal(8, "Dark2")[c(1, 2, 3, 4, 8, 5, 7)]

@@

```{r, cache = TRUE, eval = FALSE}
library(ggplot2)

@@

z$`-Pairs` = -z$Pairs
ggplot(z[z$n_sites >0, ], aes(x = `-Pairs`, y = truth)) + 
  stat_binhex(bins = 100) + 
  xlab("Estimated coefficient value") + 
  ylab("\"True\" coefficient value") + 
  stat_hline(yintercept = 0, size = 1/8) + 
  stat_vline(xintercept = 0, size = 1/8) + 
  scale_fill_gradient(low = "#F0F0F0", high = "darkblue", trans = "identity") +
  theme_bw() + 
  ggtitle("Null model (\"Pairs\")") + 
  stat_abline(intercept = 0, slope = coef(lm(z$truth ~ z$`-Pairs` + 0)))
```

@@

```{r, cache = TRUE, eval = FALSE}
rep = 12
n = 200
boots = replicate(500,
          {
            boot = lapply(
              sizes, 
              function(n){lapply(
                sample.int(max(z$rep), replace = TRUE),
                function(rep){
                  z[z$rep == rep & z$n_sites == n, ]
                }
              ) %>% bind_rows}
            ) %>% bind_rows

@@

```{r, cache = TRUE, eval = FALSE}
for(compared in c("partial correlation", "correlation", "GLM", "BayesComm", "partial BayesComm", "Pairs")){
  CI = round(
    quantile(
      boots["Markov network", ] / boots[compared, ], 
      c(.025, .975)), 
    2
  )

@@

```{r, eval = FALSE}
save(results, sizes, file = "graph_data.Rdata")
```

### 2015-10-15 Harris_2015/recovery.tex
clean up directory

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Estimated coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}"}\StringTok{True}\CharTok{\textbackslash{}"}\StringTok{ coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_hline}\NormalTok{(}\DataTypeTok{yintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_gradient}\NormalTok{(}\DataTypeTok{low =} \StringTok{"#F0F0F0"}\NormalTok{, }\DataTypeTok{high =} \StringTok{"darkblue"}\NormalTok{, }\DataTypeTok{trans =} \StringTok{"identity"}\NormalTok{) +}
\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
\StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Null model (}\CharTok{\textbackslash{}"}\StringTok{Pairs}\CharTok{\textbackslash{}"}\StringTok{)"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$Pairs +}\StringTok{ }\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
## R-squared is 1.23-1.29 times higher than from partial correlation (95% bootstrap CI)
## R-squared is 1.77-1.95 times higher than from correlation (95% bootstrap CI)
## R-squared is 1.14-1.18 times higher than from GLM (95% bootstrap CI)
## R-squared is 2.54-2.9 times higher than from BayesComm (95% bootstrap CI)
## R-squared is 1.77-1.97 times higher than from partial BayesComm (95% bootstrap CI)
## R-squared is 2.12-2.38 times higher than from Pairs (95% bootstrap CI)
\end{verbatim}

>>>>

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{=}\StringTok{ }\NormalTok{-z$Pairs}
\KeywordTok{ggplot}\NormalTok{(z[z$n_sites >}\DecValTok{0}\NormalTok{, ], }\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =} \StringTok{`}\DataTypeTok{-Pairs}\StringTok{`}\NormalTok{, }\DataTypeTok{y =} \NormalTok{truth)) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_binhex}\NormalTok{(}\DataTypeTok{bins =} \DecValTok{100}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{xlab}\NormalTok{(}\StringTok{"Estimated coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{ylab}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}"}\StringTok{True}\CharTok{\textbackslash{}"}\StringTok{ coefficient value"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_hline}\NormalTok{(}\DataTypeTok{yintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_vline}\NormalTok{(}\DataTypeTok{xintercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{size =} \DecValTok{1}\NormalTok{/}\DecValTok{8}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{scale_fill_gradient}\NormalTok{(}\DataTypeTok{low =} \StringTok{"#F0F0F0"}\NormalTok{, }\DataTypeTok{high =} \StringTok{"darkblue"}\NormalTok{, }\DataTypeTok{trans =} \StringTok{"identity"}\NormalTok{) +}
\StringTok{  }\KeywordTok{theme_bw}\NormalTok{() +}\StringTok{ }
\StringTok{  }\KeywordTok{ggtitle}\NormalTok{(}\StringTok{"Null model (}\CharTok{\textbackslash{}"}\StringTok{Pairs}\CharTok{\textbackslash{}"}\StringTok{)"}\NormalTok{) +}\StringTok{ }
\StringTok{  }\KeywordTok{stat_abline}\NormalTok{(}\DataTypeTok{intercept =} \DecValTok{0}\NormalTok{, }\DataTypeTok{slope =} \KeywordTok{coef}\NormalTok{(}\KeywordTok{lm}\NormalTok{(z$truth ~}\StringTok{ }\NormalTok{z$}\StringTok{`}\DataTypeTok{-Pairs}\StringTok{`} \NormalTok{+}\StringTok{ }\DecValTok{0}\NormalTok{)))}
\end{Highlighting}
\end{Shaded}

@@

\begin{verbatim}
## R-squared is 1.23-1.29 times higher than from partial correlation (95% bootstrap CI)
## R-squared is 1.79-1.95 times higher than from correlation (95% bootstrap CI)
## R-squared is 1.14-1.18 times higher than from GLM (95% bootstrap CI)
## R-squared is 2.72-3.28 times higher than from BayesComm (95% bootstrap CI)
## R-squared is 1.29-1.37 times higher than from partial BayesComm (95% bootstrap CI)
## R-squared is 2.12-2.37 times higher than from Pairs (95% bootstrap CI)
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{save}\NormalTok{(results, sizes, }\DataTypeTok{file =} \StringTok{"graph_data.Rdata"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

### 2015-10-15 Harris_2015/simulation.Rmd
clean up directory

---
output: pdf_document
title: Appendix [[X]]: simulating species interactions
subtitle: \"Estimating species interactions from observational data with Markov networks\"
author: David J. Harris
fontsize: 11pt
---

>>>>

---
output: pdf_document
title: Appendix 2: simulating species interactions
subtitle: \"Estimating species interactions from observational data with Markov networks\"
author: David J. Harris
fontsize: 11pt
---

### 2015-10-15 Harris_2015/three_species.Rmd
submitted to Ecology
clean up directory

# Evaluation of the three species in Figure 1

@@

    rosie = rosalia(x, trace = 0, prior = make_flat_prior(), hessian = TRUE)

    se = sqrt(diag(solve(rosie$opt$hessian)))["beta3"]
    rosalia_confident_positive = pnorm(0, mean = rosie$beta[2, 3], sd = se) < .025
    rosalia_confident_negative = pnorm(0, mean = rosie$beta[2, 3], sd = se) > .975

    c(
      cor = cor(x)[2, 3] < 0,
      partial = cor2pcor(cor(x))[2, 3] < 0,
      rosalia = rosie$beta[2,3] < 0,
      rosalia_confident_positive = rosalia_confident_positive,
      rosalia_confident_negative = rosalia_confident_negative
    )
  })

@@

```{r}
Pairs.txt = readLines("fakedata/three/Pairs.txt")

@@

```{r}
out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation)[1:3] * 100)

@@

```{r}
# percentage of Pairs runs that found statistically significant mutualism at the
# p<.05 level (two-tailed test)
mean(pnorm(pairs_z) < 0.025) * 100

>>>>

---
output: pdf_document
title: "Appendix 1: Evaluation of the three species in Figure 1"
---

@@

    rosie = rosalia(x, trace = 0, prior = make_logistic_prior(scale = 2), hessian = TRUE)

    c(
      cor = cor(x)[2, 3] < 0,
      partial = cor2pcor(cor(x))[2, 3] < 0,
      rosalia = rosie$beta[2,3],
      rosalia_se = sqrt(diag(solve(rosie$opt$hessian)))["beta3"]
    )
  })

```{r}
upper_ci_below_zero = evaluation["rosalia", ] + 
  qnorm(.975) * evaluation["rosalia_se.beta3", ] < 0
lower_ci_above_zero = evaluation["rosalia", ] + 
  qnorm(.025) * evaluation["rosalia_se.beta3", ] > 0
```

Evaluate the frequentist coverage of the confidence intervals (CIs). If the coverage is good, then approximately 95% of the confidence intervals generated on independent data sets should contain the true value (i.e., in 95% of cases, the upper CI should be above the true value of $-1$ and the lower CI should be below the true value).

```{r}
upper_ci_above_true = evaluation["rosalia", ] + 
  qnorm(.975) * evaluation["rosalia_se.beta3", ] > -1
lower_ci_below_true = evaluation["rosalia", ] + 
  qnorm(.025) * evaluation["rosalia_se.beta3", ] < -1

# Coverage is close to 95%
100 * mean(upper_ci_above_true & lower_ci_below_true)

# According to this binomial test, the confidence intervals include the
# true value a bit too often to be consistent with 95% coverage (i.e. there
# is some evidence that the confidence intervals are too conservative).
binom.test(x = sum(upper_ci_above_true & lower_ci_below_true), 
           n = length(upper_ci_above_true), 
           p = 0.95,
           conf.level = 0.95
)

@@

```{r, eval = FALSE}
Pairs.txt = readLines("fakedata/three/Pairs.txt")

@@

```{r, eval = FALSE}
out = c(pairs = mean(pairs_evaluation) * 100, rowMeans(evaluation)[1:3] * 100)

@@

```{r, eval = FALSE}
# percentage of Pairs runs that found statistically significant mutualism at the
# p<.05 level (two-tailed test)
mean(pnorm(pairs_z) < 0.025) * 100

### 2015-10-15 stochastic/chapter-3.Rmd
clean up directory

Species interactions are believed to play an important role in community structure, but detecting their influence in the co-occurrence patterns of observed communities has stymied ecologists for decades.  In recent years, several papers have suggested that ecologists could solve this long-standing problem by fitting Markov networks (undirected graphical models also known as Markov random fields) to the data by maximum likelihood or similar methods.  This approach allows ecologists to control for extraneous factors when estimating the direct interactions between a given pair of species, and previous work has shown that this kind of control is required for reliable inferences about interactions' sign and strength.  Unfortunately, the likelihood function for Markov networks quickly becomes intractable once more than about 20 species are included in the model, due to the exponentially large number of possible combinations of presences and absences. Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and attempt to explain all of the co-occurrence patterns in ecological data sets with species interactions. In this paper, I introduce "stochastic approximation" as a method for fitting these models to large ecological networks where hundreds of species and multiple abiotic factors can all affect a given species’ occurrence probability.  This approach, which converges to the maximum likelihood estimate with probability one, greatly expands the range of data sets we can analyze and the number of ecological questions we can address with these methods.

@@

To the extent that species interactions are important for community assembly, ecologists generally expect them to leave a signature on species' co-occurrence patterns.  But using that signature to infer the underlying species interactions from observational data has been much harder.  Disagreements about how to draw these inferences led to an acrimonious, decade-long argument among community ecologists [@lewin_santa_1983], where essentially all of the proposed statistical approaches were criticized for having high error rates, a poor match with the underlying ecological questions, computational infeasibility, or all three [@connor_assembly_1979; @gilpin_factors_1982; @strong_ecological_1984; @hastings_can_1987]. As documented below, ecologists have, for the most part, not solved these problems in the ensuing decades. Resolving these issues would solve a longstanding issue in community ecology and make a significant contribution to other fields, such as species distribution modeling [@kissling_towards_2012].

During the time that ecologists have struggled to identify the interactions among dozens of species, bioinformaticians have largely solved the analogous problem of estimating interactions among thousands of genes from their co-expression levels (e.g. @friedman_sparse_2008). The discrepancy is due to the fact that ecologists have focused, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients (Pollock et al. 2014; see Faisal et al. 2010 and Harris 2015 for two recent exceptions). As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (e.g. abiotic influences and indirect biotic effects; Figure 1). When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015]. While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still don't have a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.

![A hypothetical network of three species, two of which depend on an abiotic factor (annual rainfall).  **A.** The “true” network involves the caterpillar exploiting the two tree species as well as competition between the two trees.  The two trees also benefit from rainfall.  **B.** Although the two trees directly reduce one another’s occurrence probabilities via competition, they tend to occur in the same areas (with sufficient water and without too many herbivores).  An analysis that ignores these other factors could mistakenly conclude that the two trees were mutualists because of their high co-occurrence rate.  **C.** The methods presented here are designed to estimate the direct interactions between each pair of species, after accounting for other species and abiotic factors that could affect their co-occurrence rates. Note that, with observational data, the bi-directional effects of species interactions (pairs of arrows between species) need to be collapsed to a single number describing the conditional relationship between species (one double-headed arrow).](figure-1.pdf)

A number of procedures have been developed for controlling for the influence of extraneous factors and focusing on the direct relationship between a single pair of variables, i.e. for estimating the conditional relationship between them.  The most familiar of these is the partial correlation. Rather than describing the overall relationship between two variables across all conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.  @harris_estimating_2015 found that the partial correlation can do a surprisingly good job of extracting accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.

Recently, several papers have suggested that ecologists could use Markov networks (undirected graphical models also known as Markov random fields) to estimate conditional relationships from binary data by maximum entropy or maximum likelihood (@azaele_inferring_2010; @harris_estimating_2015). A Markov network defines a probability distribution over possible binary species assemblages, and its coefficients (including one coefficient describing the conditional relationship between each pair of species in the network). This approach is optimal in the sense that it produces a model that matches the observed properties of the data (i.e. the occurrence and co-occurrence rates), and that it does so with the fewest possible parameters or constraints [i.e. the Markov network has the most information entropy of any possible model satisfying its constraints; azaele_inferring_2010].  As shown in @harris_estimating_2015, Markov networks do a better job of estimating species' influences on one another better than a number of existing methods, particularly when sample sizes are low.

@@

A Markov network describes a probability distribution over possible assemblages. Since the model is a member of the exponential family it can be summarized without loss of information by its "sufficient statistics".  For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [@azaele_inferring_2010; @lee_learning_2012; @murphy_machine_2012].  One way to look at maximum likelihood estimation is that it minimizes the discrepancy between the sufficient statistics observed in the data and the expected sufficient statistics under the model [@bickel_mathematical_1977]. Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], it is possible to find the global optimum by climbing the likelihood surface. At this point, any discrepancies between the sufficient statistics of the model and of the data will have been eliminated.

In order to reduce the discrepancy between the observed and predicted sufficient statistics, the `rosalia` calculates each value exactly, averaging over all possible presence-absence combinations. This approach is infeasible for larger networks, as the number of possibilities grows exponentially with the size of the network. Stochastic approximation [@robbins_stochastic_1951; @salakhutdinov_efficient_2012] is a procedure that allows us to approximate the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees.  The procedure iterates through the following three steps as many times as needed (50,000 for these analyses):

1: simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.

2: subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient

3: Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.

The specific procedure used here had the following implementation details:

The simulations in Step 1 used Gibbs sampling to generate examples of landscapes based on the model’s current parameter estimates. While the simulated landscapes produced by Gibbs sampling are serially autocorrelated, statisticians have shown that this merely slows convergence to the maximum likelihood estimate rather than preventing it altogether [@younes_convergence_1999; @salakhutdinov_efficient_2012].

The approximate likelihood gradients in Step 2 match the ones from @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations. These gradients were augmented with a momentum term [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 2.0 (for environmental responses) or 0.5 (for pairwise interactions).

The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = \alpha_0 1000/(999 + t)$, which satisfies the criteria in @powell_approximate_2007. The scaling factor, $\alpha_0$, was 1.0 for species’ intercepts and environmental responses. Because of the enormous number of pairwise interaction coefficients, I set $\alpha_0$ to 0.1 for the $\beta$ parameters.

@@

Each species was assigned two two sets of coefficients.  The coefficients determining species’ responses to the environment were each drawn from standard normal distributions.  The coefficients describing species’ pairwise interactions were drawn from a mixture of normals (see Appendix) so that most interactions were weak and negative, but a few strong positive and interactions also occurred.

@@

![Stochastic approximation improves the model over time. **A.** With small networks where the exact maximum likelihood estimates are known, stochastic approximation reaches the global optimum very quickly. Within a few minutes, the squared deviation between the approximate and exact estimates disappears. **B.** With larger networks, where the maximum likelihood estimate cannot be calculated exactly, stochastic approximation converges to an apparent optimum that explains most (but not all) of the variation in the "true" parameter values.](convergence.pdf)

>>>>

Species interactions are believed to play an important role in community structure, but detecting their influence in the co-occurrence patterns of observed communities has stymied ecologists for decades.  While Markov networks (undirected graphical models also known as Markov random fields) represent a promising approach to this long-standing problem by isolating direct interactions from indirect ones, the methods ecologists have suggested for fitting these models are limited to small communities with about 20 species or fewer. Additionally, the methods that have been proposed so far do not account for environmental heterogeneity, and thus attempt to explain all of the co-occurrence patterns in ecological data sets with species interactions. In this paper, I introduce stochastic approximation as an alternative method for fitting these models that addresses both of these problems, making it feasible to use Markov networksin cases where each species responds to multiple abiotic factors and hundreds of competitors or mutualists. While stochastic approximation does introduce some sampling noise during the optimization process, it still converges to the maximum likelihood estimate for species' interaction coefficients with probability one.

@@

To the extent that species interactions are important for community assembly, ecologists generally expect them to leave a signature on species' co-occurrence patterns.  Using that signature to infer the underlying species interactions from observational data has been much harder, however.  Disagreements about how to draw these inferences led to an acrimonious, decade-long argument among community ecologists [@lewin_santa_1983], where essentially all of the proposed statistical approaches were criticized for having high error rates, a poor match with the underlying ecological questions, computational infeasibility, or all three [@connor_assembly_1979; @gilpin_factors_1982; @strong_ecological_1984; @hastings_can_1987]. As documented below, ecologists have, for the most part, not solved these problems in the ensuing decades. Resolving these issues would solve a longstanding issue in community ecology and make a significant contribution to other fields, such as species distribution modeling [@kissling_towards_2012].

During the time that ecologists have struggled to identify the interactions among dozens of species, bioinformaticians have largely solved the analogous problem of estimating interactions among thousands of genes from their co-expression levels (e.g. @friedman_sparse_2008). The discrepancy is due to the fact that ecologists have chosen to focus, almost exclusively, on overall co-occurrence rates [@connor_assembly_1979; @gotelli_empirical_2009; @gilpin_factors_1982; @veech_probabilistic_2013] or on closely-related values such as correlation coefficients (Pollock et al. 2014; see Faisal et al. 2010 and Harris 2015 for two recent exceptions). As ecologists know, however, the correlation between two species will often reflect other factors beyond their direct pairwise interactions (e.g. abiotic influences and indirect biotic effects; Figure 1). When these factors are not accounted for, this reliance on overall co-occurrence rates can reliably lead to incorrect inferences [@harris_estimating_2015]. While some methods have been proposed for incorporating a small number of specific factors (such as geographic or environmental dissimilarity) into ecologists' null models at a time [@lessard_strong_2011], we still lack a good way to scale these approaches up for cases with many biotic and abiotic factors acting simultaneously.

![A hypothetical network of three species, two of which depend on an abiotic factor (annual rainfall).  **A.** The “true” network involves the caterpillar exploiting the two tree species as well as competition between the two trees.  The two trees also benefit from rainfall.  **B.** Although the two trees directly reduce one another’s occurrence probabilities via competition, they tend to occur in the same areas (with sufficient water and without too many herbivores).  An analysis that ignores these other factors could mistakenly conclude that the two trees were mutualists because of their high co-occurrence rate (circled "+" sign).  **C.** The methods presented here are designed to estimate the direct interactions between each pair of species, after accounting for other species and abiotic factors that could affect their co-occurrence rates (note that the circled interaction between the trees has the correct sign). Note that, with observational data, the bi-directional effects of species interactions (pairs of arrows between species) need to be collapsed to a single number describing the conditional relationship between species (one double-headed arrow).](figure-1.pdf)

A number of procedures have been developed for controlling for the influence of extraneous factors and focusing on the direct relationship between a single pair of variables, i.e. for estimating the conditional relationship between them.  The most familiar of these is the partial correlation. Rather than describing the overall relationship between two variables across all conditions, partial correlations (like regression coefficients) describe the portion of the relationship that remains after the other variables in the data have been accounted for.  @harris_estimating_2015 found that the partial correlation can do extract accurate information about pairwise species interactions from binary co-occurrence data in some circumstances, but the fact that this approach assumes multivariate Gaussian data makes it less appealing.

Recently, several papers have suggested that ecologists could use Markov networks (undirected graphical models also known as Markov random fields) to estimate conditional relationships from binary data by maximum entropy or maximum likelihood (@azaele_inferring_2010; @harris_estimating_2015). A Markov network defines a probability distribution over possible binary species assemblages, and its coefficients (including one coefficient describing the conditional relationship between each pair of species in the network). This approach is optimal in the sense that it produces a model that matches the observed properties of the data (i.e. the occurrence and co-occurrence rates), and that it does so with the fewest possible parameters or constraints [i.e. the Markov network has the most information entropy of any possible model satisfying its constraints; @azaele_inferring_2010].  As shown in @harris_estimating_2015, Markov networks do a better job of estimating species' influences on one another better than a number of existing methods, particularly when sample sizes are low.

@@

A Markov network describes a probability distribution over possible assemblages. Since the model is a member of the exponential family it can be summarized without loss of information by its "sufficient statistics".  For the model presented here, the sufficient statistics include: the number of occurrences for each species, the number of co-occurrences between each species, and the cross-product between the species and environment matrices [@azaele_inferring_2010; @lee_learning_2012; @murphy_machine_2012].  Finding the maximum likelihood estimate for the model parameters is equivalent to minimizing the discrepancy between the values of the data's sufficient statistics and the expected sufficient statistics under the model  [@bickel_mathematical_1977]. Because fully-observed Markov networks, such as the ones analyzed here, have unimodal likelihood functions [@murphy_machine_2012], it is possible to find the global optimum by iteratively reducing the discrepancies between the sufficient statistics of the model and of the data to zero.

In order to reduce the discrepancy between the observed and predicted sufficient statistics, the `rosalia` calculates each value exactly, averaging over all possible presence-absence combinations. Stochastic approximation [@robbins_stochastic_1951; @salakhutdinov_efficient_2012] instead estimates the expected values of the sufficient statistics by averaging over a more manageable number of simulated assemblages during each model-fitting iteration, while still retaining maximum likelihood convergence guarantees.  The procedure iterates through the following three steps as many times as needed (50,000 for these analyses; see Appendix for annotated R code):

1. Simulate a set of assemblages from the current model parameters and calculate sufficient statistics for the sample.

2. Subtract the simulated sufficient statistics from the observed ones to calculate the approximate likelihood gradient

3. Adjust the model parameters to climb the approximate gradient, using a schedule of step sizes that satisfies the criteria in Chapter 6 of @powell_approximate_2007.

Here, the simulations in Step 1 used Gibbs sampling to generate examples of landscapes based on the model’s current parameter estimates. While the simulated landscapes produced by Gibbs sampling are serially autocorrelated, statisticians have shown that this merely slows convergence to the maximum likelihood estimate rather than preventing it altogether [@younes_convergence_1999; @salakhutdinov_efficient_2012].

The approximate likelihood gradients in Step 2 match the ones from @harris_estimating_2015, except that they are averaged over a set of Monte Carlo samples rather than over all possible presence-absence combinations. These gradients were augmented with a momentum term [@hinton_practical_2012] and by regularizers based on a logistic prior with location 0 and scale 2.0 (for environmental responses) or 0.5 (for pairwise interactions). The step size parameter in Step 3, $\alpha_t$, decreased after each iteration according to a generalized harmonic sequence, $\alpha_t = 1000/(999 + t)$, which satisfies the criteria in @powell_approximate_2007.

@@

Each species was assigned two sets of coefficients.  The coefficients determining species’ responses to the environment were each drawn from standard normal distributions.  The coefficients describing species’ pairwise interactions were drawn from a mixture of normals (see Appendix) so that most interactions were weak and negative, but a few strong positive and interactions also occurred.

@@

![Stochastic approximation improves the model over time. **A.** With small networks where the exact maximum likelihood estimates are known, stochastic approximation approaches the global optimum very quickly (note the log scale of the y axis). Within a few seconds, the mean squared deviation between the approximate and exact estimates drops to less than 0.01; in less than a minute, it drops below 0.0001. For comparison, the `rosalia` package took about six minutes to find the maximum likelihood estimate. **B.** With larger networks, where the maximum likelihood estimate cannot be calculated exactly, stochastic approximation converges to an apparent optimum that explains most (but not all) of the variation in the "true" parameter values.](convergence.pdf)
